Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2017-06-20 2017-Xudong-SMT/submit-ICTAC2017.tex
Section 3 et 4

\newcommand{\Pred}{\symb{Pred}}
\newcommand{\Post}{\symb{Post}}
%\usepackage[math]{cellspace}
%\setlength\cellspacetoplimit{ 37pt}
%\setlength\cellspacebottomlimit{18pt}

@@

We choose the expression \texttt{P>>(acc(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using pNets.

This pNet is composed of: a toplevel node PN1 encoding the Enable
operator, with its controller (pLTS) C1; a hole representing the
process variable P; another node PN2 representing the Prefix
operator, with its controller C2, and its hole Q.

@@

Now look more specifically on the top pNet node PN1. In addition to
its controller it has 2 sub-pNets: the hole P and the sub-pNet PN2.
Three synchronization vectors are defining the possible
synchronizations between the actions of the subnets.
 The first
vector, transmitting an action $a_1$ of P, can only occur if $a_1$ is not
a $\delta$, the vector concrete syntax is typically ``$< l, a_1 , \_ >
-> a_1$'', in which ``$\_$'' means that the corresponding subnet is
not involved in this synchronization, and the ``l'' action is related
to the corresponding transition of the controller C1: this action is
available only when the state variable $s_0$ is equal to 0.
The second vector composes the
actions from the subnets and gives out a synchronized action
$\underline{\delta(x)}$, where the underlined action stands for an
internal (already synchronized) action, that cannot be further
synchronized at upper levels. Synchronized action is a straightforward
generalization of the notion of internal actions, that will be
convenient for observing internal events during model-checking. Here
the transition occurs only when process
$P$ performs a $\delta(x)$ action, transferring the control to $Q$,
and sending a value $x$ at the same time. This value must be
``accepted''  through an $acc(x)$ action before $Q$ takes the
control. The last vector acts as the first one, but transmitting the
action $a_2$ of Q.  
Similarly the node PN2 has two synchronization vectors to encode
the behavior of the Prefix operator.

@@

It should be easy to see that this pNet system encodes properly the
behavioral semantics of the corresponding algebra expression. We remark
that it is built in a systematic structural way, one pNet node
encoding each operator. It is parameterized both at the level of
actions variables (occurring within the controllers and the
synchronization vectors), and at the level of process parameters
(holes).

@@

\QIN{[How about reduce the example of OT? But we still need keep the proof tree example.]}

The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes transitions of several LTSs with
action of some holes, the transition occurs if some predicates hold, and can involve a 
set of state modifications.

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} over a set  $(S_i,s_{0 i}, \rightarrow_i)^{i\in
	I}$ of LTSs, a
	set $J$ of holes with sorts $Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred, \Post}
	{s \xrightarrow {v}s'}
	\end{mathpar}
	Where $s, s'\in\mathcal{S}$ and for all
        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
	LTS $(S_i,s_{0 i}, \rightarrow_i)$, and $\xrightarrow{b_j}_j$
        is a transition of the hole $j$, for any action $b_j$ in the
        sort $Sort_j$. \Pred\ is a predicate
	over the different variables of the
	terms, labels, and states $s_i$, $b_j$, $s$, $v$. \Post\ is a set of equations that 
	hold \emph{after the open transition}, they are represented as a substitution of the 
	form $\{x_k\gets e_k\}^{k\in K}$ 
	where $x_k$ are variables of $s'$, $s'_i$, and $e_k$ are expressions over the other 
	variables of the open transition.
\end{definition}

@@

\TODO{Relate to the algorithm, or even move to section 5}
When building an Open Automaton as the semantics of a pNet, its
\emph{states}, and the shape of the predicates in its
\emph{transitions} have a specific structure:

@@

\paragraph{Predicates:}
Let
$\mylangle\overline{\pNet},\overline{S},\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We define a
predicate $\Pred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\[\Pred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Leftrightarrow
\begin{array}{l}
\exists {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.\, SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
\\~~\land
\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{array}\]
In any other case (if the action families do not match or if there is no valuation of
variables such that the above formula can be ensured) the predicate is undefined.

@@

%\QIN{[ change the example OT here ]}
\begin{example}\emph{An open-transition.}
  \label{OT:lotos-composed}

The \texttt{LOTOS} pNet of Fig. \ref{schema:lotos-pnet} has 2
controllers and 2 holes. We will show its full open Automaton in
section \ref{section:full-result}. We give one of its transition here:

@@

The semantic of the open pNets is already defined as an open automaton in \cite{henrio:Forte2016} using open transitions to present the transitions of its global states.
\begin{definition}[Operational semantics of open pNets]
  \label{def:operationalSemantics}
  The semantics of a pNet $p$ is an open automaton $A = <Leaves(p),J,\mathcal{S}, s_0,
  \mathcal{T}>$ where:
  \begin{itemize}
    \item $J$ is the indices of the holes: $Holes(p)= H_j^{j\in J}$.
    %  \item $\overline{L}^L = Leaves(p), \overline{H}^J = Holes(p)$
    \item $\overline{\mathcal{S}} = States(p)$ and $s_0 = InitState(p)$
    \item $\mathcal{T}$ is the smallest set of open transitions
    satisfying the rules below:
  \end{itemize}

@@

  The second rule deals with pNet nodes: for each possible
  synchronization vector applicable to the rule subject, the premisses
  include one {\em open transition} for each sub-pNet involved, one possible
  {\em action} for each Hole involved, and the predicate relating these
  with the resulting action of the vector.
  A key to understand this rule is that the open transitions are
  expressed in terms of the leaves and holes of the pNet structure,
  i.e. a flatten view of the pNet: e.g. $L$ is the index set of the
  Leaves, $L_k$ the index set of the leaves of one subnet, so all $L_k$
  are disjoint subsets of $L$. Thus the states in the open transitions,
  at each level, are tuples including states of all the
  leaves of the pNet, not only those involved in the chosen
  synchronization vector.

@@

  \noindent
    $\inferrule
    {k\!\in\! K \\ SV\!\!= \!clone(SV_k) \!=\! \alpha_m^{m \in I_k\uplus J_k} \!\to\! 
    \alpha'_k, G_k \\
      Leaves(p) \!=\! \pLTS_l^{l\in L} \\     
      \forall m\in I_k.
      \pNet_m \models
      \inferrule*[myfraction=\reddottedrule]
      {\{s_{i}\xrightarrow{a_{i}}_i s_{i}'\}^{i\in I_m^\prime},
      \{\xrightarrow{b_{j}}_j\}^{j\in J'_m}, \Pred_m, \Post_m}
      {\ostate{s_{i}^{i \in L_m}} \xrightarrow {v_m}
        \ostate{s_{i}^{\prime\ i \in L_m}}}
      %\land
      %Leaves(\pNet_m) = \overline{pLTS}^{L_k})
      \\
      I' = \biguplus_{m\in I_k}\!\! I_m'
      \\ J' = \biguplus_{m\in I_k}\!\! J'_m \uplus J_k  \\
      \Pred = \bigwedge_{m\in I_k}\!\! \Pred_m \land
      \Pred(SV,v_m^{m\in I_k},b_j^{j\in J_k},v)\\
      \forall j\!\!\in\!\! J_k. {\tt 
        fresh}(b_j) \\ {\tt fresh}(v) \\ 
      \forall i\in
      L\backslash I'.\,s'_i=s_i 
    }
    {p = \mylangle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\myrangle
      \models
      {\inferrule*[myfraction=\reddottedrule]
        {\{s_i\xrightarrow{a_i}_i s_i^{\prime}\}^{i\in I^\prime},
        \{\xrightarrow{b_j}_j\}^{j\in J^\prime}, \Pred, \uplus_{m\in I_k} 
        \Post_m}
        {\ostate{s_i^{i\in L}} \xrightarrow {v}
          \ostate{s_i^{\prime i\in L}}}
      }
    }
    $

@@

\TODO{To be move  by Xudong}

@@

\TODO{To be move / replace by Xudong}
%\subsection{Computing and using open automata}
%In this section we present an algorithm to construct the open
%automaton representing the behavior of an open pNet, and show
%under which conditions this automaton is finite.
%
%\begin{alg}[behavioral semantics of open pNets: Sketch]
%This is a standard residual algorithm over a set of open-automaton
%states, but where transitions are open transitions
%constructively ``proven'' by deduction trees.
%
%1) Start with a set of unexplored states containing the initial state
%of the automaton, and an empty set of explored states.
%
%2) While there are unexplored states:
%
%2a) pick one state from the unexplored set and add it to the explored
%set. From this state
%build all possible deduction trees by application of the structural
%rules Tr1 and Tr2, using all applicable combinations
%of synchronization vectors.
%
%2b) For each deduction trees, extract the resulting
%open-transition.
%Its predicate is checked (using Z3) for  satisfiability. If it is
%unsatisfiable, the resulting OT is discarded. This will minimize the
%number of resulting transitions. 
%
%2c) For each open-transition,  add the transition in the outgoing transitions of
%the current state, and add the
%resulting state in the unexplored set if it is not already in the
%explored set.
%
%3) Finally a reachability check could potentially
%prune the state space.
%
%\end{alg}

@@

To have some practical interest, it is important to know when this
algorithm terminates. The following theorem shows that an open-pNet
with finite synchronization sets, finitely many leaves and
holes, and each pLTS at leaves having a finite number of states and
(symbolic) transitions, has a finite automaton. The proof can be found
in \cite{henrio:Forte2016}.

\begin{theorem}[Finiteness of open-automata.]\\
Given an open pNet $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $pLTS_i^{i\in L}$ and holes $Hole_j^{j\in
  J}$, if the sets $L$ and $J$ are finite, if the synchronization vectors of all pNets 
  included in  $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ 
  are finite, and if
$\forall i \in L.\, finite{(states(pLTS_i))} \text{ and } pLTS_i$
has a finite number of state variables, then Algorithm 1 terminates
and produces an open automaton 
$\mathcal{T}$ with finitely many states and transitions.

@@

% $$ finite{(L)} \land finite{(J)} \land \forall i \in L finite{(\{s_i\})}
%   \to finite{(\mathcal{T})}$$
\end{theorem}

>>>>

\newcommand{\Pred}{\symb{Pred}}
\newcommand{\MkPred}{\symb{MkPred}}
\newcommand{\Post}{\symb{Post}}
%\usepackage[math]{cellspace}
%\setlength\cellspacetoplimit{ 37pt}
%\setlength\cellspacebottomlimit{18pt}

@@

We choose the expression \texttt{P>>(acc(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using pNets. 
This pNet is composed of: a toplevel node PN1 encoding the Enable
operator, with its controller (pLTS) C1; a hole representing the
process variable P; another node PN2 representing the Prefix
operator, with its controller C2, and its hole Q.

@@

Now consider the top pNet node PN1. In addition to
its controller it has 2 sub-pNets: the hole P and the sub-pNet PN2.
Three synchronization vectors define the possible
synchronizations between the actions of its subnets.
The first vector ``$< l, a_1 , \_ > -> a_1$'', transmitting an action
$a_1$ of P, can only occur if $a_1$ is not  
a $\delta$, here ``$\_$'' means that the corresponding subnet is
not involved in this synchronization, and the ``l'' action is related
to the corresponding transition of the controller C1: this action is
available only when the state variable $s_0$ is equal to 0.
The second synchronisation vector composes an
action $\delta(x)$ from $P$ with an $acc(x)$ from $Q$, transmitting a
value $x$ before $Q$ takes the control; 
The vector result $\underline{\delta(x)}$ is an
internal action, called ``synchronized'' that cannot be further
synchronized at upper levels. Synchronized actions is a straightforward
generalization of the notion of internal actions, that will be
convenient for observing internal events during model-checking.
The last vector acts is to the first one.  
Similarly the node PN2 has two synchronization vectors to encode
the behavior of the Prefix operator.

It should be easy to see that this pNet system encodes properly the
behavioral semantics of the corresponding algebra expression. We remark
that it is built in a systematic structural way, one pNet node
encoding each operator.
%% It is parameterized both at the level of
%% actions variables (occurring within the controllers and the
%% synchronization vectors), and at the level of process parameters
%% (holes).

\TODO{I'm thinking about moving this second example to section 5...}

@@

@@

The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes transitions
of several LTSs with the
actions of some holes; the transition occurs if some predicates hold,
and can involve a set of state modifications.

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} over a set  $(S_i,s_{0 i}, \rightarrow_i)^{i\in
	I}$ of LTSs, a
	set $J$ of holes with sorts $Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred, \Post}
	{s \xrightarrow {v}s'}
	\end{mathpar}
	Where $s, s'\in\mathcal{S}$ and for all
        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
	LTS $(S_i,s_{0 i}, \rightarrow_i)$, and $\xrightarrow{b_j}_j$
        is a transition of the hole $j$, for any action $b_j$ in the
        sort $Sort_j$. \Pred\ is a predicate
	over the variables of the terms, labels, and states $s_i$,
        $b_j$, $s$, $v$. \Post\ is a set of equations that  
	hold \emph{after the open transition}, represented as a
        substitution $\{x_k\gets e_k\}^{k\in K}$  
	where $x_k$ are variables of $s'$, $s'_i$, and $e_k$ are
        expressions over the other variables of the open transition.
\end{definition}

@@

% \TODO{Relate to the algorithm, or even move to section 5}
When building an Open Automaton as the semantics of a pNet, its
\emph{states}, and the shape of the predicates in its
\emph{transitions} have a specific structure:

@@

\paragraph{Predicates:}
Let
$\mylangle\overline{\pNet},\overline{S},\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\[\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Leftrightarrow
\begin{array}{l}
\exists {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.\, SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
\\~~\land
\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{array}\]
In any other case (if the action families do not match or if there is no valuation of
variables such that the above formula can be ensured) the predicate is undefined.

@@

\begin{example}\emph{An open-transition.}
  \label{OT:lotos-composed}

The \texttt{LOTOS} pNet of Fig. \ref{schema:lotos-pnet} has 2
controllers and 2 holes. We will show its full open Automaton in
section \ref{section:full-result}. We give one of its transitions here:

@@

% The semantic of the open pNets is already defined as an open automaton
% in \cite{henrio:Forte2016} using open transitions to present the
% transitions of its global states.

\begin{definition}[Operational semantics of open pNets]
  \label{def:operationalSemantics}
  The semantics of a pNet $p$ is an open automaton $A = <Leaves(p),J,\mathcal{S}, s_0,
  \mathcal{T}>$ where:
  \begin{itemize}
    \item $J$ is the indices of the holes: $Holes(p)= H_j^{j\in J}$.
    %  \item $\overline{L}^L = Leaves(p), \overline{H}^J = Holes(p)$
    \item $\overline{\mathcal{S}} = States(p)$ and $s_0 = InitState(p)$
    \item $\mathcal{T}$ is the smallest set of open transitions
    satisfying the rules below:
  \end{itemize}

@@

  The second rule ({\bf Tr2}) deals with pNet nodes: for each possible
  synchronization vector applicable to the rule subject, the premisses
  include one {\em open transition} for each sub-pNet involved, one possible
  {\em action} for each Hole involved, and the predicate relating these
  with the resulting action of the vector.
  A key to understand this rule is that the open transitions are
  expressed in terms of the leaves and holes of the pNet structure,
  i.e. a flatten view of the pNet: e.g. $L$ is the index set of the
  Leaves, $L_k$ the index set of the leaves of one subnet, so all $L_k$
  are disjoint subsets of $L$. % Thus the states in the open transitions,
%   at each level, are tuples including states of all the
%   leaves of the pNet, not only those involved in the chosen
%  synchronization vector.

@@

  \noindent
    $\inferrule
    {k\!\in\! K \\ SV\!\!= \!clone(SV_k) \!=\! \alpha_m^{m \in I_k\uplus J_k} \!\to\! 
    \alpha'_k, G_k \\
      Leaves(p) \!=\! \pLTS_l^{l\in L} \\     
      \forall m\in I_k.
      \pNet_m \models
      \inferrule*[myfraction=\reddottedrule]
      {\{s_{i}\xrightarrow{a_{i}}_i s_{i}'\}^{i\in I_m^\prime},
      \{\xrightarrow{b_{j}}_j\}^{j\in J'_m}, \Pred_m, \Post_m}
      {\ostate{s_{i}^{i \in L_m}} \xrightarrow {v_m}
        \ostate{s_{i}^{\prime\ i \in L_m}}}
      %\land
      %Leaves(\pNet_m) = \overline{pLTS}^{L_k})
      \\
      I' = \biguplus_{m\in I_k}\!\! I_m'
      \\ J' = \biguplus_{m\in I_k}\!\! J'_m \uplus J_k  \\
      \Pred = \bigwedge_{m\in I_k}\!\! \Pred_m \land
      \MkPred(SV,v_m^{m\in I_k},b_j^{j\in J_k},v)\\
      \forall j\!\!\in\!\! J_k. {\tt 
        fresh}(b_j) \\ {\tt fresh}(v) \\ 
      \forall i\in
      L\backslash I'.\,s'_i=s_i 
    }
    {p = \mylangle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\myrangle
      \models
      {\inferrule*[myfraction=\reddottedrule]
        {\{s_i\xrightarrow{a_i}_i s_i^{\prime}\}^{i\in I^\prime},
        \{\xrightarrow{b_j}_j\}^{j\in J^\prime}, \Pred, \uplus_{m\in I_k} 
        \Post_m}
        {\ostate{s_i^{i\in L}} \xrightarrow {v}
          \ostate{s_i^{\prime i\in L}}}
      }
    }
    $

@@

% \TODO{To be move  by Xudong}

@@

% \TODO{To be move / replace by Xudong}

%\subsection{Computing and using open automata}
%In this section we present an algorithm to construct the open
%automaton representing the behavior of an open pNet, and show
%under which conditions this automaton is finite.
%
%\begin{alg}[behavioral semantics of open pNets: Sketch]
%This is a standard residual algorithm over a set of open-automaton
%states, but where transitions are open transitions
%constructively ``proven'' by deduction trees.
%
%1) Start with a set of unexplored states containing the initial state
%of the automaton, and an empty set of explored states.
%
%2) While there are unexplored states:
%
%2a) pick one state from the unexplored set and add it to the explored
%set. From this state
%build all possible deduction trees by application of the structural
%rules Tr1 and Tr2, using all applicable combinations
%of synchronization vectors.
%
%2b) For each deduction trees, extract the resulting
%open-transition.
%Its predicate is checked (using Z3) for  satisfiability. If it is
%unsatisfiable, the resulting OT is discarded. This will minimize the
%number of resulting transitions. 
%
%2c) For each open-transition,  add the transition in the outgoing transitions of
%the current state, and add the
%resulting state in the unexplored set if it is not already in the
%explored set.
%
%3) Finally a reachability check could potentially
%prune the state space.
%
%\end{alg}

@@

%% To have some practical interest, it is important to know when this
%% algorithm terminates. The following theorem shows that an open-pNet
%% with finite synchronization sets, finitely many leaves and
%% holes, and each pLTS at leaves having a finite number of states and
%% (symbolic) transitions, has a finite automaton. The proof can be found
%% in \cite{henrio:Forte2016}.

%% \begin{theorem}[Finiteness of open-automata.]\\
%% Given an open pNet $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $pLTS_i^{i\in L}$ and holes $Hole_j^{j\in
%%   J}$, if the sets $L$ and $J$ are finite, if the synchronization vectors of all pNets 
%%   included in  $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ 
%%   are finite, and if
%% $\forall i \in L.\, finite{(states(pLTS_i))} \text{ and } pLTS_i$
%% has a finite number of state variables, then Algorithm 1 terminates
%% and produces an open automaton 
%% $\mathcal{T}$ with finitely many states and transitions.

@@

% $$ finite{(L)} \land finite{(J)} \land \forall i \in L finite{(\{s_i\})}
%   \to finite{(\mathcal{T})}$$
% \end{theorem}
